### 2021-03-05 paper.Rmd
added bullet point for arguments
working on initial example with plot

source("misc-functions.R")
set.seed(2207) # 2207

@@

A different approach to anticlustering directly works with feature based data 
and does not require a transformation into pairwise dissimilarities. Instead, 
the sum of the squared Euclidean distances between elements and the cluster they 
are assigned to is minimized [@brusco2006branch], most prominently using the 
k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion 
simultaneously maximizes between-cluster separation and within-cluster 
homogeneity [@aloise2009np]. Späth [-@spath1986] and Valev [-@valev1998; 
-@valev1983] independently recognized that by optimizing the reversed k-means 
objective that is, maximizing the sum of distances between elements and cluster 
centroids -- a data set is partitioned into similar groups, thereby reversing 
the logic of classical clustering. They coined the term anticlustering for this 
purpose. K-means anticlustering maximizes the sum of squared Euclidean distances 
from each element to the center of the cluster to which it belongs. As 
@spath1986 noted, k-means anticlustering thereby directly minimizes differences 
with regard to the means of the numeric variables between groups. Hence, groups 
only become similar with regard to the mean of the distribution of the numeric 
attributes, which may perform poorly as a measure of overall similarity; the 
spread of the data is not targeted and may differ between sets. @papenberg2020 
showed that maximizing the diversity (based on the pairwise Euclidean distances) 
is better suited to minimize difference with regard to both the mean and the 
variance of the data; it was shown that even an entirely random split usually 
resulted in more similar standard deviations as k-means solely focuses on the 
mean. When requiring that groups are overall similar to each other, this focus 
may be misguided as similar means can be obtained even when the underlying 
distributions are different [@anscombe1973].

@@

Note that a minor problem arises when ignoring the weights during optimizing the 
k-plus objective. Since their computation involves squaring, the squared 
distances $y_{ij}$ have much larger values, on average, than the raw values 
$x_{ij}$. Therefore, the $SSE_{Var}$ criterion naturally receives greater weight 
when computing the k-plus criterion than the standard $SSE$ criterion. This 
imbalance is unwarranted because optimizing similarity with regard to variances 
is not more important than optimizing similarity with regard to means. A simple 
solution for this issue is to standardize all values $x_{ij}$ and $y_{ij}$ to 
the same scale before computing $SSE_{kplus}$, e.g. via $z$-standardization.

@@

if (!file.exists(path)) {
  # Generate data and all partitions
  N <- 14
  K <- 2
  partitions <- generate_partitions(N, K)
  features <- rnorm(N, 15, 3)

  # Compute objectives for each partition
  all_objectives_mean <- sapply(
    partitions,
    FUN = diff_mean,
    features = features
  )

  all_objectives_var <- sapply(
    partitions,
    FUN = diff_vars,
    features = features
  )

  # For each partition: Compute SSE objective, and SSE(VAR)
  # SSE
  all_objectives_kMeans <- sapply(
    partitions,
    FUN = var_objective,
    features = features
  )

  # SSE(VAR)
  all_objectives_kVar <- sapply(
    partitions,
    FUN = var_objective,
    features = squared_from_mean(as.matrix(features))
  )

  df <- data.frame(
    diff_mean = all_objectives_mean,
    diff_var  = all_objectives_var,
    kmeans = all_objectives_kMeans,
    kvar = all_objectives_kVar
  )

@@

Optimizing a bicriterion objective such as $SSE_{kplus}$ may not succeed as one 
may hope. The ideal outcome would be to find a partition where both the means 
and the variances are very similar between groups. However, research on 
multi-objective optimization generally shows that cases are rare where a single 
solution satisfies all criteria to a satisfactory degree. Usually, a tradeoff is 
inevitable. It is an open question whether k-plus anticlustering is a expection 
from this rule. Next, a motivating example illustrates that in anticlustering, 
aiming to simultaneously optimize similarity of means and variances is not only 
desirable, but may actually be feasible. This notion is later confirmed when 
applying k-plus anticlustering in simulation studies and examples using real 
data.

For this initial example, I created 14 data points from a univariate normal 
distribution ($M = 15$, $SD = 3$) and generated all `r n_partitions(14, 2)` 
possible ways to partition the 14 data points into $K = 2$ equal-sized groups. 
For each partition, the criteria SSE and $SSE_{Var}$ were computed. Figure 1 
shows the scatterplot of these criteria across partitions. Note that for both 
criteria, a larger value is better, i.e., indicates higher similarity with 
regard to means and variances, respectively. Ideally, to obtain overall 
between-group similarity, a partition is chosen that minimizes the difference 
with regard to both means and variances (see the triangular symbol). However, 
optimal k-means anticlustering would create two groups that are not particularly 
similar with regard to their variance. The partition that is represented by the 
squared symbol has the overall best highest SSE, but is outperformed by many 
other partitions with regard to $SSE_{Var}$. The diamond symbol represents 
that partition that where the two groups have the least difference in 
variances.

Next: use the partition that optimizes $SSE_{kplus}$!

@@

# Illustrate Pareto set
#pareto_pts <- pareto_set(df[, c("kmeans", "kvar")])
#points(df[pareto_pts, c("kmeans", "kvar")], pch = 19)

# Illustrate best partition wrt difference in means
points(
  df$kmeans[which.max(df$kmeans)],
  df$kvar[which.max(df$kmeans)],
  cex = 1.2, pch = 22, bg = "white", lwd = 1.7
)

# Illustrate best partition wrt difference in variance
points(
  df$kmeans[which.max(df$kvar)],
  df$kvar[which.max(df$kvar)],
  cex = 1.2, pch = 24, bg = "white", lwd = 1.7
)

df$kplus <- rowSums(df[, c("kmeans", "kvar")])

# Illustrate best partition wrt simililarity combined similarity / variance
points(
  df$kmeans[which.max(df$kplus)],
  df$kvar[which.max(df$kplus)],
  cex = 1.2, pch = 23, bg = "white", lwd = 1.7
)
```

>>>>

source("misc-functions.R")

@@

A different approach to anticlustering directly works with feature based data 
and does not require a transformation into pairwise dissimilarities. Instead, 
the sum of the squared Euclidean distances between elements and the cluster they 
are assigned to is minimized [@brusco2006branch], most prominently using the 
k-means heuristic [@jain2010; @steinley2006]. Minimizing the k-means criterion 
simultaneously maximizes between-cluster separation and within-cluster 
homogeneity [@aloise2009np]. Späth [-@spath1986] and Valev [-@valev1998; 
-@valev1983] independently recognized that by optimizing the reversed k-means 
objective that is, maximizing the sum of distances between elements and cluster 
centroids -- a data set is partitioned into similar groups, thereby reversing 
the logic of classical clustering. They coined the term anticlustering for this 
purpose. K-means anticlustering maximizes the sum of squared Euclidean distances 
from each element to the center of the cluster to which it belongs. As 
@spath1986 noted, k-means anticlustering thereby directly minimizes differences 
with regard to the means of the numeric variables between groups. Hence, groups 
only become similar with regard to the mean of the distribution of the numeric 
attributes, which may perform poorly as a measure of overall similarity; the 
spread of the data is not targeted and may differ between sets. @papenberg2020 
showed that maximizing the diversity (based on the pairwise Euclidean distances) 
is better suited to minimize difference with regard to both the mean and the 
variance of the data; it was# Variance objective function with changed order of the arguments 
# (can be called by sapply() etc.)
var_objective <- function(clusters, features) {
  anticlust::variance_objective(features, clusters)
}
 shown that even an entirely random split usually 
resulted in more similar standard deviations as k-means solely focuses on the 
mean. When requiring that groups are overall similar to each other, this focus 
may be misguided as similar means can be obtained even when the underlying 
distributions are different [@anscombe1973].

@@

Note that a minor problem arises when ignoring the weights during optimizing the 
k-plus objective. Since their computation involves squaring, the squared 
distances $y_{ij}$ have much larger values, on average, than the raw values 
$x_{ij}$ (also see the axes in Figure 1). Therefore, the $SSE_{Var}$ criterion 
naturally receives greater weight when computing the k-plus criterion than the 
standard $SSE$ criterion. This imbalance is unwarranted because optimizing 
similarity with regard to variances is not more important than optimizing 
similarity with regard to means. A simple solution for this issue is to 
standardize all values $x_{ij}$ and $y_{ij}$ to the same scale before computing 
$SSE_{kplus}$, e.g. via $z$-standardization.

@@

if (!file.exists(path)) {
  set.seed(2207) # 2207
  # Generate data and all partitions
  N <- 14
  K <- 2
  partitions <- generate_partitions(N, K)
  features <- rnorm(N, 15, 3)

  # Compute objectives for each partition
  # For each partition: Compute SSE objective, and SSE(VAR)
  # 1. SSE
  all_objectives_kMeans <- sapply(
    partitions,
    FUN = var_objective,
    features = features
  )

  squared_distances_from_mean <- squared_from_mean(as.matrix(features))

  # 2. SSE(VAR)
  all_objectives_kVar <- sapply(
    partitions,
    FUN = var_objective,
    features = squared_distances_from_mean
  )

  # 3. SSE(KPLUS)
  # Use standardization to compute k-plus criterion!
  all_objectives_kPlus <- sapply(
    partitions,
    FUN = var_objective,
    features = scale(cbind(features, squared_distances_from_mean))
  )

  df <- data.frame(
    kmeans = all_objectives_kMeans,
    kvar = all_objectives_kVar,
    kplus = all_objectives_kPlus
  )

@@

When a bicriterion objective such as $SSE_{kplus}$, the ideal outcome would be 
to find a partition where both the means and the variances are very similar 
between groups, i.e., a which partition where both SSE and $SSE_{Var}$ are close 
to the optimum value. However, research on multi-objective optimization 
generally shows that cases are rare where a single solution satisfies all 
criteria to a satisfactory degree. Usually, a tradeoff is necessary. For 
anticlustering, however, a motivating example illustrates that aiming to 
simultaneously optimize similarity of means and variances is not only desirable, 
but may actually be feasible.

@@

# Illustrate best partition wrt difference in means (i.e., SSE)
points(
  df$kmeans[which.max(df$kmeans)],
  df$kvar[which.max(df$kmeans)],
  cex = 1.2, pch = 22, bg = "white", lwd = 1.7
)

# Illustrate best partition wrt difference in variance (i.e., SSE(VAR))
points(
  df$kmeans[which.max(df$kvar)],
  df$kvar[which.max(df$kvar)],
  cex = 1.2, pch = 24, bg = "white", lwd = 1.7
)

# Illustrate best partition wrt k-plus criterion
points(
  df$kmeans[which.max(df$kplus)],
  df$kvar[which.max(df$kplus)],
  cex = 1.2, pch = 23, bg = "white", lwd = 1.7
)
```

For this initial example, I created 14 data points from a univariate normal 
distribution ($M = 15$, $SD = 3$) and generated all `r n_partitions(14, 2)` 
possible ways to partition the 14 data points into $K = 2$ equal-sized groups. 
Figure 1 shows the scatterplot of the criteria SSE and $SSE_{Var}$ across 
partitions. Note that for both criteria, a larger value is better, i.e., 
indicates higher similarity with regard to means and variances, respectively. 
Therefore, to obtain overall between-group similarity, a partition should be 
chosen that minimizes the difference with regard to both means and variances, 
i.e., a partition that is depicted in the upper right corner of the scatterplot. 
However, optimal k-means anticlustering would create two groups that are not 
particularly similar with regard to their variance. The partition that is 
represented by the squared symbol has the overall best highest SSE, but is 
outperformed by many other partitions with regard to $SSE_{Var}$. Similarly, the 
partition that maximizes $SSE_{Var}$ (the triangular symbol in Figure 1) is not 
well-suited to optimize SSE at the same time. The diamond symbol represents the 
partition that maximizes $SSE_{kplus}$, i.e., sum of SSE and $SSE_{Var}$. This 
partition is located exactly in the right upper corner and yields `r 
df$kmeans[which.max(df$kplus)] / max(df$kmeans) * 100`% of the global maximum 
SSE, and `r df$kvar[which.max(df$kplus)] / max(df$kvar) * 100`% of the global 
maximum value of $SSE_{Var}$. Hence, this initial example shows that maximizing 
the combined k-plus criterion may well be suited to optimize both similarity 
with regard to means as well as variances between groups. Later, I will show 
that this result also holds for larger data sets, more groups, and more 
features.
